---
title: "Cytokine analysis of orthopedic patients"
format:
  html:
    toc: true
    self-contained: TRUE
editor: visual
theme: darkly
fig-height: 7
fig-width: 7 
---

```{r}
#| label: load-packages
#| include: false

library(tidyverse)
```

## Load data

```{r}
#| label: load data

pg <- read.table("pg_ml.tsv",sep="\t",header=TRUE,row.names=1)
pg[1:6,1:8]
str(pg)

pg <- t(pg)
pg[1:6,1:8]

```

## My custom analysis

Basic analysis.

```{r}
#| label: basics

dim(pg)

colcount <- apply(pg,2, function(x) {
  length(which(!is.na(x)))
})

head(colcount)
barplot(colcount,main="num cytokines detected per sample")

rowcount <- apply(pg,1, function(x) {
  length(which(!is.na(x)))
})

head(rowcount)
barplot(rowcount,main="number of samples with a cytokine detected")

rowcount <- rowcount[order(rowcount)]
barplot(head(rowcount,20),horiz = TRUE, las=1,cex.names = 0.8,xlab="num samples",main="detection rate")

```

## MDS plot

Running an MDS plot.

It shows some variability, in particular 4015, 2007, 1007 and some others are not clustered with the bulk of samples.

```{r}
#| label: mds1

mds <- cmdscale(dist(t(pg)))

cols <- sapply(strsplit(colnames(pg),""),"[[",1)
cols <- gsub("2","green",cols)
cols <- gsub("1","lightblue",cols)
cols <- gsub("3","pink",cols)
cols <- gsub("4","darkgray",cols)


plot(mds, xlab="Coordinate 1", ylab="Coordinate 2", 
  col=cols, cex=4 , pch=19, main="MDS plot all samples (not transformed)",
  bty="n")

text(mds,labels = colnames(pg))

```

Now log transform the data and repeat.

It certainly looks better, as there is distinct clustering of the blue samples which is distinct from the other samples.

```{r}
#| label: mds2

lpg <- log(pg)
mds <- cmdscale(dist(t(lpg)))

cols <- sapply(strsplit(colnames(pg),""),"[[",1)
cols <- gsub("2","green",cols)
cols <- gsub("1","lightblue",cols)
cols <- gsub("3","pink",cols)
cols <- gsub("4","darkgray",cols)


plot(mds, xlab="Coordinate 1", ylab="Coordinate 2", 
  col=cols, cex=4 , pch=19, main="MDS plot all samples (not transformed)",
  bty="n")

text(mds,labels = colnames(pg))

```

## References

## Session information

It is always a good idea to record information about the environment, which will help reproducibility.

```{r}
#| label: sessioninfo
#| warning: false
sessionInfo()
```
